Verifying Spatial Properties of Array Computations

### 2017-04-18 oopsla17/algorithms.tex
various edits from sections 2-6

\subsection{Union Normal Form}\label{subsec:union-normal-form}

\noindent
We need to address the problem of representation in the model for index schemes.
In each case, the objects are sets and union and intersection are the
combinators. However, they are often infinite sets due to \bz{} in Cartesian
product. This happens when constant indices of array terms are converted to
\bz{} in subscript schemes and when infinite interval is generated in interval
schemes. This calls for a finite (and ideally compact) representation for
indices.

First recall \Cref{lem:vector-intersect}, which states that intersection of
indexing schemes is still an indexing scheme. When combined with
\Cref{lem:interval-identities}, we can extend this closure to interval schemes.
Subscripts schemes do not enjoy the same propery as intersection of two
singleton sets might easily be empty, which disqualifies it from being a
subscript index. In the following sections we will not rely on intersection of
subscript indices, so that is not a problem.

Union, however, is not as flexible as intersection. \Cref{lem:vector-union} does
not form a closure for indexing schemes in general. For this reason, we
represent collection of all forms of indexing schemes as unevaluated unions.
Since the model forms a distributive lattice as established in
\Cref{prop:region-lattice}, the intersections can be pushed in using the
distributive law leaving only unions. We further exploit associativity of union
to put unions into a cons tree making the unions effectively a non-empty list.
We do not, however, attempt to canonicalise the representation. Later on
\Cref{lem:vector-union} and index scheme subset are used to reduce the size of
union normal form.

@@

\newcommand{\neigh}{\textsf{neigh}}

The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathit{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable.% where we use the predicate
%\arrayTy{} to classify variables of array type.

@@

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{sec:study}, we establish dimensonality
99\% of all array variables that we can assign a specification is X, and 99\% of
number of array terms that flows into such an assignment is Y. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

@@

The preceding source code analysis converts the concrete syntax of
array subscripts into a set of subscript schemes in
$\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
into sets of index schemes. We denote sets of index schemes by
$\mathcal{S}$, and initial set of subscript schemes from the analysis
as $\mathcal{S}_0$. For our example:
%
\begin{equation*}
\mathcal{S}_0 = \{\{0\} \times \{0\}, \{-1\} \times \{0\},
\{0\} \times \{-1\}, \{0\} \times \{1\}, \{1\} \times \{0\}\}
\end{equation*}
%
%For each array variable $a \in \mathsf{dom}(U)$, the algorithm
%constructs a set of $n$-dimensional rectangles covering all contiguous
%groups of schemes in $U(a)$.
%
We explain the algorithm in two parts. First, we show how
\emph{contiguous} index schemes are coalesced into a smaller set of
index schemes while remaining in union normal form. Then, we explain
how this smaller unioned indexing schemes are converted to
specifications.

>>>>

\begin{example}
\label{exm:checking}
We demonstrate the analysis and modelise procedures over the
following example kernel which computes the mean value
of a five-point stencil:
%%
\begin{minted}{fortran}
x = a(i-1, j)+a(i+1, j); y = a(i, j-1)+a(i, j+1)
b(i, j) = (a(i, j) + x + y) / 5.0
\end{minted}
\end{example}

@@

\newcommand{\neigh}{\textsf{neigh}}
\noindent
The analysis builds on standard program analyses:
%
\textbf{(1)} basic blocks (CFG);
\textbf{(2)} induction variables per basic block;
\textbf{(3)} (interprocedural) data-flow analysis, providing a \emph{flows to}
  graph (as shorthand, the function
  $\mathit{flowTo}$ is used, implicitly parameterised by this graph,
  mapping an expression to the set of all expressions
  with forwards data-flow to this expression, based on the transitive
  closure of the flows graph);
\textbf{(4)} type information per variable.% where we use the predicate
%\arrayTy{} to classify variables of array type.

The analyse then traverse the control-flow graph of a program and,
traverses statements inside of loops in reverse. The right-hand side
of an assignment statement is classified on whether it has array
subscripts which are neighbourhood offsets (or potential absolute in
some dimensions). These are then converted into our model domain by
the interpretation $\interp{-}^{\mathit{aterm}}$
(~\Cref{{defn:subscript}) and grouped into a finite
  map from array variables to set of subscript schemes.

We denote sets of index/subscripts schemes by
$\mathcal{S}$, For our example, the initial set of subscript schemes from the analysis
is given by $\mathcal{S}_0$ as:
%
\begin{equation*}
\mathcal{S}_0 = \{\{0\} \times \{0\}, \{-1\} \times \{0\},
\{0\} \times \{-1\}, \{0\} \times \{1\}, \{1\} \times \{0\}\}
\end{equation*}

@@

Although satisfiability is super-exponential with the length of the formula in
the worst case~\cite{fischer1974super}, performance of consistency checking
using satisfiability is fast in practice as the length of the formula is linearly
related with two factors: (1) dimensionality times number of regions connected
with \texttt{+} and (2) dimensionality times the number of array terms flowing
into an assignment assignment. In \Cref{subsec:additional-data}, we
established that  99.6\% of array computations have 
dimensionality of at most four, and 97\% of array computations
involve at most 4 array subscript terms. In
\Cref{sec:evaluation}, we show the number of regions that are connected with
\texttt{+} is Z in 99\% of the cases.

@@

The preceding source code analysis converts the concrete syntax of
array subscripts into a set of subscript schemes in
$\mathcal{P}(\bz)^N$. These subscripts are then coalesced iteratively
into sets of index schemes.

%
%For each array variable $a \in \mathsf{dom}(U)$, the algorithm
%constructs a set of $n$-dimensional rectangles covering all contiguous
%groups of schemes in $U(a)$.
%
We explain the algorithm in two parts. First, we show how
\emph{contiguous} index schemes are coalesced into a smaller set of
index schemes while remaining in union normal form. Then, we explain
how this smaller unioned indexing schemes are converted to
specifications.

### 2017-04-18 oopsla17/evaluation.tex
'non-trivial' count
various edits from sections 2-6

One of the initial goals of CamFort was to provide tools that enable
optimisation and refactoring without changing behaviour, and the
stencil specification feature was designed with that in mind. But
while we were perusing real-world source bases, we realised that there
were also many cases where CamFort could help with refactoring
precisely because they required changes in behaviour. For example, a
relatively common change observed in the logs of revision control for
various projects is the refactoring of array indices. Either
re-ordering, adding or deleting indices. Any of these, if not
perfectly propagated throughout all uses, could result in unexpected
outcomes. However, CamFort will pick up the difference during its
stencil specification checking. Ideally, an experienced programmer may
take advantage of the region variable feature of CamFort in order to
minimise the number of specifications that need changing and gain the
full time-saving advantage. But even if not, a quick search for the
specifications can have them updated quickly and then any code that is
not updated will trigger an error when checked.

>>>>

One of the initial goals of CamFort was to provide tools that enable
optimisation and refactoring without changing behaviour, and the
array specification feature was designed with that in mind. But
while we were perusing real-world source bases, we realised that there
were also many cases where CamFort could help with refactoring
precisely because they required changes in behaviour. For example, a
relatively common change observed in the logs of revision control for
our corpus is the refactoring of array indices. Either
re-ordering (swapping dimensions), or adding or deleting indices. Any of these, if not
perfectly propagated throughout all uses, could result in unexpected
outcomes. However, CamFort will pick up the difference during its
stencil specification checking. Ideally, an experienced programmer may
take advantage of the region variable feature of CamFort in order to
minimise the number of specifications that need changing and gain the
full time-saving advantage. But even if not, a quick search for the
specifications can have them updated quickly and then any code that is
not updated will trigger an error when checked.

### 2017-04-18 oopsla17/introduction.tex
various edits from sections 2-6

Our approach is not targeting a class of bugs that can be detected
automatically (\emph{push-button verification}). Instead, array
indexing bugs must be identified relative to a specification of the
intended access pattern.  Nevertheless, we report on instances of code
revisions from the commit histories of our corpus where a
correct specification would have either spotted an error (which was
latter correct) or would have assisted in an array refactoring task
(\Cref{sec:case-studies}).

>>>>

Our approach does not target a class of bugs that can be detected
automatically (\emph{push-button verification}). Instead, array
indexing bugs must be identified relative to a specification of the
intended access pattern.  Nevertheless, we report on instances of code
revisions from the commit histories of our corpus where a
correct specification would have spotted an error (which was
latter corrected) or would have assisted in refactoring 
(\Cref{sec:case-studies}).

### 2017-04-18 oopsla17/lang.tex
various edits from sections 2-6

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
   \nonterm{rconst} \; \mid \; \nonterm{region} \, \term{+}
  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \;
    \nonterm{region} \\
\nonterm{rconst} ::= & \stencil{p}{\dimTy}{}{} \; \\
\mid \; & \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
\mid \; & \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
\mid \; & \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\[0.25em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

\subsection{Specification syntax}
\label{subsec:syntax}

@@

Region specifications can be assigned to region variables
(\emph{rvar}) via region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection
convolution~\cite{davis1975survey} can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common patterns, such as the five-point
pattern, as the regions can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).
The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

>>>>

\begin{figure}[t]
\begin{align*}
\def\arraystretch{1}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{$\mathbb{N}_{>0}$}
\newcommand{\depthRange}{$\mathbb{N}_{>0}$}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \; \mid \; \term{access} \;
                        \nonterm{spec} \; \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \mid \;
   \nonterm{rconst} \; \mid \; \nonterm{region} \, \term{+}
  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \;
    \nonterm{region} \\
\nonterm{rconst} ::= & \stencil{p}{\dimTy}{}{} \; \\
\mid \; & \stencil{f}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
\mid \; & \stencil{b}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\
\mid \; & \stencil{c}{\dimTy}{\depthRange}{\;[, \nonpointed]} \\[0.25em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\end{figure}

%\subsection{Specification syntax}
%\label{subsec:syntax}

@@

Region specifications can be assigned to region variables
(\emph{rvar}) via region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection
convolution~\cite{davis1975survey} can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common patterns, such as the five-point
pattern, as the regions can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).
The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the
array subscripts were repeated. This modifier provides a way to
rule out any accidental repetition of array subscripts.
The notion is similar to that of linear types~\cite{wadler1990linear}, where a value must be used
exactly once. We opt for the more informative and easily understood name
\texttt{readOnce}. This modifier is optional, so it need not
be present even if the stencil is linear.

@@

\paragraph{Relativisation}

The data access pattern of a stencil is understood relative to the
subscript on the left-hand side. If the left-hand side of a stencil
has a neighbourhood subscript with non-zero offsets, then the
right-hand side is understood relative to this shifted origin. For example:
%%
\begin{minted}{fortran}
do i = 1, n
   != stencil (backward(depth=2, dim=1, nonpointed))*(pointed(dim=2)) :: b
   a(i+1, j) = b(i, j) + b(i-1, j)
\end{minted}
%%
Note that the specification for dimension 1 is to a depth of 2, since
the left-hand side is shifted forward by 1.

\paragraph{Access specifications}

Specifications can also be given to array computations which are not
stencils (\ie{}, do not have a left-hand side which is an array
subscript). For example, for a reduction:
%%
\begin{minted}{fortran}
do i = 1, n
  != access, pointed(dim=1) :: a
  r = max(a(i), r)
end do
\end{minted}
%%
The main difference between \fortran{stencil} and \fortran{access} is
that there is no left-hand side with which to check consistency or
relativise against.

### 2017-04-18 oopsla17/model.tex
various edits from sections 2-6

\begin{defn}
  Recall array subscript terms of the form $a(\bar{e})$ from
  \Cref{def:array-subs}. We interpret these terms with the partial
  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
  all indices are either constant or neighbourhood indices as defined in
  \Cref{def:neighbour-ix}.
%
  \begin{align*}
    \interp{a(\bar{e})}^{\mathit{aterm}} =
      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
  %
    \textit{subscript}(e) = \begin{cases}
      \{ c \} & e \equiv i \pm c \\
      \bz & e \; \mbox{is constant}
    \end{cases}
  \end{align*}
\end{defn}
%
Note that this produce subscript schemes (a form of index scheme), but
they are not members of \textit{Interval} or $\textit{Region}_N$. In
the next section, we use this interpretation on array subscripts in
the static analysis producer preceding checking and inference.

>>>>

\begin{defn}
  \label{defn:subscript}
  Recall array subscript terms of the form $a(\bar{e})$ from
  \Cref{def:array-subs}. We interpret these terms with the partial
  interpretation $\interp{-}^{\mathit{aterm}} : \textit{array-term}
  \prightarrow{} \mathcal{P}(\bz{})^N$. The interpretation is defined when
  all indices are either constant or neighbourhood indices as defined in
  \Cref{def:neighbour-ix}.
%
  \begin{align*}
    \interp{a(\bar{e})}^{\mathit{aterm}} =
      \prod_{1 \leq i \leq N} \mathit{subscript}(\bar{e}_i) & &
  %
    \textit{subscript}(e) = \begin{cases}
      \{ c \} & e \equiv i \pm c \\
      \bz & e \; \mbox{is absolute}
    \end{cases}
  \end{align*}
\end{defn}
%
Note that this produce subscript schemes (a form of index scheme), but
they are not members of \textit{Interval} or $\textit{Region}_N$. In
the next section, we use this interpretation on array subscripts in
the static analysis producer preceding checking and inference.

\subsection{Union Normal Form}\label{subsec:union-normal-form}

\noindent
We lastly address a problem of \emph{representation} for index schemes
in the model. Index schemes may incorporate the infinite set \bz{} in
their components, \eg{} when a dimension is unconstrained or when an
absolute index of array term is interpreted. In order to provide a
finite checking procedure, we require a finite and compact
representation for indices which accounts for this.

Recall \Cref{lem:vector-intersect} states that the
intersection of indexing schemes is still an indexing scheme. When
combined with \Cref{lem:interval-identities} (union and intersection
on holed intervals), we can extend this closure to interval schemes.
Subscripts schemes do not enjoy the same propery as intersection of
two singleton sets might easily be empty, which disqualifies it from
being a subscript index. In the following sections we will not take
intersection of subscript indices.

Union, however, is not as flexible as
intersection. \Cref{lem:vector-union} does not form a closure for
indexing schemes in general. For this reason, we represent sets of
indexing schemes as unevaluated union terms, or \emph{union normal
  form}.  Since the model forms a distributive lattice as established
in \Cref{prop:region-lattice}, the intersections can be pushed inwards
using the distributive law leaving only unions at the outer-level. We
further exploit associativity of union to put unions into a cons-tree
making the unions effectively a non-empty list.  We do not, however,
attempt to canonicalise the representation. Later on
\Cref{lem:vector-union} and index scheme subset are used to reduce the
size of union normal form.
